Type-Safe Modular Parsing

### 2017-06-05 SLE_17/sections/Inheritance.tex
Fit into 12 pages
Section 3

\section{Parsing OO ASTs with Multiple Inheritance}\label{sec:inheritance}

@@

Part of the modular parsing problem is how to obtain an extensible parser. It
is natural to make use of OO ASTs because adding new data constructs is cheap for
them. Hence we have used OO traits and inheritance to represent the AST in the last
section. Furthermore, we would like to write extensible parsing code on extensions of
a grammar. That is to say, new extensions would not require modification on the existing code,
and we can even reuse the old code.

@@

Unfortunately, this fails to parse some expressions like \inlinecode{"1 + x"}, which are obviously valid in the new grammar.
The reason is that \inlinecode{pAdd} makes two recursive calls to parse sub-expressions, by using \inlinecode{pExpr}, which
covers both cases in the old grammar. But the newly added case \inlinecode{pVar} is not observed by the recursive \inlinecode{pExpr},
hence the parser does not work as expected. It is possible to build the correct parser by replacing the recursive call in \inlinecode{pAdd} with \inlinecode{pVarExpr}.
However, modification on existing code sacrifices separate compilation, and hence breaks semantic modularity.

@@

In the above code, we use the \lstinline{super} keyword to refer to the old \inlinecode{pExpr}.
With multiple inheritance, Scala has a special syntax
\lstinline{super[T].x}, called \textit{static super reference}. It refers to the type or
method \inlinecode{x} in the specified parent trait \inlinecode{T}. With this we can reuse
the old parsers even if they have the same name.

>>>>

\section{\hspace{-10pt}OO AST Parsing with Multiple Inheritance}\label{sec:inheritance}

@@

Part of the modular parsing problem is how to obtain an extensible parser. It
is natural to make use of OO ASTs because adding new data constructs is cheap for
them. Hence we have used OO traits and inheritance to represent the AST in the last
section. Furthermore, we would like to write extensible parsing code on extensions of
a grammar. That is to say, new extensions would not require modifying the existing code,
and we can even reuse the old code.

@@

Unfortunately, this fails to parse some expressions like \inlinecode{"1 + x"}, which are obviously valid in the new grammar.
The reason is that \inlinecode{pAdd} makes two recursive calls to parse sub-expressions, by using \inlinecode{pExpr}, which
covers both cases in the old grammar. But the newly added case \inlinecode{pVar} is not observed by the recursive \inlinecode{pExpr},
hence the parser does not work as expected. It is possible to build the correct parser by replacing the recursive call in \inlinecode{pAdd} with \inlinecode{pVarExpr}.
However, modifying existing code sacrifices separate compilation, and hence breaks semantic modularity.

@@

In the above code, we use the \lstinline{super} keyword to refer to the old \inlinecode{pExpr}.
With multiple inheritance, Scala has a special syntax
\lstinline{super[T].x}, called \textit{static super reference}. It refers to the name
\inlinecode{x} in the specified parent trait \inlinecode{T}. With this we can reuse
the old parsers even if they have the same name.
\bruno{Independent extensibility cannot be explained like this. 
We need to see some code here! You need to expand the explanation. 
I suggest instead cutting on the ``wrong'' attempt in the beggining of this section, and instead 
have some brief discussion in the \emph{correct} code where we need to be careful.}

### 2017-06-05 SLE_17/sections/PackratParsing.tex
Fit into 12 pages
Section 3

Traditional alternative composition will only find the first parser that succeeds on a prefix of the input,
in spite that subsequent parsers may parse the whole input. In contrast with the parser above, \inlinecode{"parseAdd <|> parseInt"} works as expected with the two cases swapped.

In this case, reordering the alternatives ensures that
the \emph{longest match} is picked among the possible results. However, manual reordering for the longest match is inconvenient, and worst still, it is essentially non-modular. When the grammar is extended with new rules, programmers are supposed to \emph{manually} adjust
the order of parsers, by rewriting previously written code.

\paragraph{Backtracking} The need for backtracking can also be problematic
in a modular setting. Consider a grammar that includes \inlinecode{"import..from"}.
It is extended with an \inlinecode{"import..as"} case:

@@

Similarly, this violates a modular setting because it also requires a global view of the full grammar.
Hence the worst case should be considered that all alternatives may share common prefixes with future cases. In that case we need to backtrack for all the branches. To avoid failures in the future, we have to add \inlinecode{try} everywhere. However this results in the worst-case exponential time
complexity.

@@

It is worth mentioning that the left-recursive grammar above is well supported without extra code.
 The longest match composition is also employed by using the combinator \lstinline{|||}. Furthermore, it does not suffer from the backtracking problem, as the memoization technique of Packrat parsing guarantees reasonable efficiency.

>>>>

Traditional alternative composition will only find the first parser that succeeds on a prefix of the input,
even if subsequent parsers may parse the whole input. In contrast to the previous parser, \inlinecode{"parseAdd <|> parseInt"} works as expected with because the two cases are swapped.
In this case, reordering the alternatives ensures that
the \emph{longest match} is picked among the possible results. However, manual reordering for the longest match is inconvenient, and worst still, it is essentially non-modular. When the grammar is extended with new rules, programmers should \emph{manually} adjust
the order of parsers, by rewriting previously written code.

\paragraph{Backtracking} The need for backtracking can also be problematic
in a modular setting. Consider a grammar that includes \inlinecode{"import..from"}, 
and is extended with an \inlinecode{"import..as"} case:

@@

Similarly, this violates a modular setting because it also requires a global view of the full grammar.
Hence the worst case where all alternatives may share common prefixes with future cases
should always be antecipated. Therefore we need to backtrack for all the branches. To avoid failures in the future, we have to add \inlinecode{try} everywhere. However this results in the worst-case exponential time
complexity.

@@

It is worth mentioning that the left-recursive grammar above is well supported without extra code.
 The longest match composition is also employed by using the combinator \lstinline{|||}. Furthermore, the parser does not suffer from the backtracking problem, as the memoization technique of Packrat parsing guarantees reasonable efficiency.
